##
Red Wine EDA by Will Tays
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

#install.packages('GGally')
#install.packages('gridExtra')
#install.packages('corrplot')

library(ggplot2)
library(reshape2)
library(GGally)
library(gridExtra)
library(corrplot)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('winequality-red.csv', sep = ';')

```

#### 
I chose the red wine dataset for a few personal reasons. I enjoy drinking
red wine, and I'm a chemist. Howerver, I've never considered studying the 
chemistry underlying wine. This project allowed me the time to dig into how a 
wine is made and all the chemical components that must be balanced to create a 
quality wine. 

# Data attributes and structure

```{r echo=FALSE, Univariate_Plots}
head(df)
dim(df)
names(df)
str(df)
summary(df)

```

# Univariate Plots

```{r echo=FALSE}
# histogram matrix for all columns
meltData <- melt(df)

ggplot(aes(x=value), data=meltData) +
  geom_histogram() +
  facet_wrap(~variable, scales = 'free_x')

```

####
Generating a matrix of histograms for all columns provides a global view of the 
chemical attributes measured in the red wine contained in this data set. These 
attributes can be used to explore whether a single, or combination of attributes 
can impact the quality rating given by wine experts. Overall, each attribute
seems to be normally distributed with the exception of maybe citric acid, 
free sulfur dioxide, and alcohol.  Although normally distributed, most tend to 
lean or skew to the right. 

We'll zoom in on each plot to get a more detailed look at each histogram. 



```{r echo=FALSE}
# function to streamline histogram creation
histplot <- function(variable, ...) {
 ggplot(data=df, aes_q(as.name(variable))) +
    geom_histogram(..., color='black') + 
    ggtitle(variable)
}
col.summary <- function(variable){
  print(summary(df[[variable]]))
}
hist.summary <- function(variable, ...) {
  col.summary(variable)
  histplot(variable, ...)
}

```

```{r echo=FALSE}
# histograms for the acids in wine
# fixed.acidity is right-skewed distributed
# volatile.acidity is almost normally distributed at 0.52, but also right-skewed
# citric.acid is right-skewed as a large amount of wines in the dataset contain virtually none
grid.arrange(hist.summary('fixed.acidity', bins=25),
             hist.summary('volatile.acidity', bins=20) +
               xlim(0, 1.2), 
             hist.summary('citric.acid', bins=15) )
```

#### 
Fixed acidity (mainly tartaric acid) can determine how tart a wine may taste. 
We can see that a majority of the wines in the dataset contain less than 12 g/L. 
A high acidity wines "tartness" may be balanced with more sugar to sweeten the 
profile. 

Volatile acids (mainly acetic acid) evaporate easier than the fixed acids in
wine. High concentrations of volatile acids can casue unpleasent flavor and 
aroma profiles. 

The histogram for citric acid is results in a distrubion which is not normally
distributed. I increased and decreased the binwidth and failed to see a normal
distribution. Citric acids are typically added after fermentation as doing so 
before may risk microbes to convert it to acetic acid, which is undesirable in 
the wine making process. It makes sense looking at the histgram for citric acid 
to see very lowconcentrations, most less than mean at 0.271 g/L. 

```{r echo=FALSE}
# salt and sugars 
# residual.sugar content is very low which reflects here as 3rd quartile of data is less than 2.60
grid.arrange(hist.summary('residual.sugar', bins=20) +
      scale_x_continuous(breaks = seq(0, 15, 2.5)) +
      xlim(0,10),
      hist.summary('chlorides', bins=25) +
    xlim(0, 0.25))


```

#### 
The relatively low residual sugar in these wines suggest that the flavor 
profiles for the bulk of this dataset is dry (or not sweet). According to the 
info given with the data, wines greater than 45 g/L are considered sweet. In our
dataset, the max sugar content for a wine is 15.5 g/L. If we limit the x-axis 
(0, 10) to remove some of the outliers, the histrogram still skews heavily to 
the right.  

The histrogram for chrlorides can be adjusted to remove values greater than
0.25 to get a better look at the distribution. Salt content in wine aims to be 
very minimal as wine is not meant to be salty. The salt profile comes from dry 
soils and even the skin of the grapes. 

```{r echo=FALSE}
# Sulfur species in wine
# free.SO is right-skewed
# total.SO is right-skewed 
# sulphates look to be normally distributed if we removed some ourliers
grid.arrange(hist.summary('free.sulfur.dioxide', bins=20),
              hist.summary('total.sulfur.dioxide', bins=20) +
            xlim(0,200),
            hist.summary('sulphates', bins=20) + 
            xlim(0,1.5))

```

#### 
The frequencies for free and total sulfur dioxide similar may be due to total
SO2 being the sum of free SO2 and bound forms of SO2. Although not naturally 
occuring, sulphates are added and can contritbute to SO2 levels. 


```{r echo=FALSE}
# density looks to be normally distributed, centered at 0.9968
hist.summary('density', bins=20)

```

#### 
Density looks to be normally distributed with almost identical mean (0.9967) 
and median (0.9968).

```{r echo=FALSE} 
# pH is normally distributed, centered at 3.31
hist.summary('pH', bins=15)

```

#### 
The histogram for pH shows a normal distribution centered at a median of 3.30.
The mean pH is 3.31. 
Most wines tend to lie between 3-4 on the pH scale, making them acidic. This 
makes sense as controlling the acidity of wine is a very important during wine
making. 


```{r echo=FALSE}
# alcohol content seems to skew-right slightly
hist.summary('alcohol', bins=15)

```

#### 
Like citric acid, alcohol content seems to have a different distrubiton.
Adjusting the binwidth showed that alcohol leans right, and is almost like the 
second half of a normal distribution curve. The median and mean are almost
equivalent (10.20, 10.42), but a good frequency of wines have alcohol content 
higher than the mean/median, ranging up to a max of 14.9 percent/volume. 

```{r echo=FALSE}
# create a bar chart to visualize the quality ratings
# quality rating seems to center heavily between 5 and 6 
ggplot(aes(x=quality), data=df) +
  geom_bar() +
  scale_x_continuous(breaks = seq(3, 8, 1)) +
  ggtitle('Quality Ratings breakdown for red wines')

```

#### 
The quality rating, which in this case visuzlizes best with a bar chart shows
that most of the ratings lie between 5-6 . The entire usable scale is 0-10, but 
our dataset only exhibits wine in the 3-8 qualilty range. 

# Univariate Analysis

####
The red wine data contains 1599 observations(rows) and 12 variables (columns). 
Each variable represents a chemical attribute of red wine along with a 'quality' 
rating.  Thus, the data is very favorable for comparing quality with specific 
attributes such as acidity or alcohol. Hisotgrams for each attribute that most 
are distributed normally with skew (mostly to the right). Some unusual 
distributions were exhibited with citric acid and alcohol content. With some 
adjustments to binwidth and x-axis to remove outliers, the distribution still
failed to show they were normal. 


# Bivariate Plots Section

```{r echo=FALSE, fig.height=10, fig.width=10,Bivariate_Plots}

ggpairs(df)

#ggsave('wine matrix.pdf')

```

```{r echo=FALSE}
# generating a correlation plot 
df.cor <- cor(df)

corrplot(df.cor, method = 'shade', type = 'upper', order = 'original',
         tl.srt = 90, addCoef.col = 'black', tl.col = 'red', 
         number.cex = 0.6, tl.cex = 0.8, diag = FALSE)

```

```{r echo=FALSE}
# list of correlations comparing available attributes with quality
cor(df, df$quality)

```

### 
Using the GGally and corrplot library to create matrix plot, we can visualize a 
lot of details in a single plot to explore any points of interest. Since this 
data looks to find what attributes contribute to quality, we look for positive 
correlations when comparing to quality. For example, we may assume that alcohol 
content may contribute to higher quality ratings and find that, while not 
particularly high (corr = 0.476), it is the strongest factor in unlocking a good 
quality rating. 
Though there are some negative correlations associated with quality, volatile
acid content has the largest impact on lowering quality rating. 

```{r echo=FALSE}
# boxplot of alcohol vs quality
ggplot(aes(x=quality, y=alcohol, group=quality), data = df) +
  geom_boxplot() + 
  scale_x_continuous(breaks = seq(3, 8,  1)) +
  ggtitle('Wines with more alcohol lead to better quality')

```

#### 
Even if there is overlap, we see that the median alcohol rises as quality
rises with the exception at rating 5. 



```{r echo=FALSE}
# comparing alcohol vs quality
q1 <- ggplot(aes(x=quality, y=alcohol), data=df) +
  geom_jitter(alpha=0.3) +
  scale_x_continuous(breaks = seq(3, 8, 1)) +
  geom_smooth(method = 'lm')

# comparing sulphates vs quality
q2 <- ggplot(aes(x = quality, y=sulphates), data = df) +
  geom_jitter(alpha=0.3) +
  geom_smooth(method = 'lm')

# comparing citric acidity vs quality
q3 <- ggplot(aes(x=quality, y=citric.acid), data = df) +
  geom_jitter(alpha=0.3) +
  geom_smooth(method = 'lm')

grid.arrange(q1, q2, q3, heights=1)

```

#### 
Comparing the three attributes with the highest correlations to quality
rating, we see that alcohol content has a stronger effect on quality. The 
higer the alcohol content, the higher the quality rating given. This may make 
some sense in a social aspect as the effects of alcohol may lead to a more 
favorable mood. Citric acid is not naturally found in wine, and can be added in
small amounts to give a freshness and additional flavor to the wine. This also
seems to line up with a better quality rating as the citric acid adds freshness
lead and potentially better tasting wine. 

Sulphates are also an additive which are used to prevent microbial growth and 
oxidation. They may also, though minimally,  contribute to both species of 
sulfur dioxide (SO2) which, at high enough concentrations, can change the aroma 
and flavor profile. 

```{r echo=FALSE}
# mean alcohol content per quality rating
by(df$alcohol, factor(df$quality), summary)

# mean citric acid conentrations per quality rating
by(df$citric.acid, factor(df$quality), summary)

# mean sulphate concentrations per quality rating
by(df$sulphates, factor(df$quality), summary)

```

#### 
Interestingly, while examining the effects per quality rating, alcohol content
at rating 5 actually had a lower mean than at rating 4. This may be due a number
of factors including a higher frequency of wines rated at 5 comparted to 4. 
However, it's a noteworthy finding considering I assumed that the mean alcohol
concentration would continually increase with higher quality ratings. 



```{r echo=FALSE}
# volatile acidity vs quality has a negative impact
ggplot(aes(x=quality, y=volatile.acidity, group=quality), data = df) +
  geom_boxplot(alpha=0.4) +
  scale_x_continuous(breaks = seq(3, 8, 1)) +
  ggtitle('High concentrations of acetic acids negatively impact quality')

```

####
Not surprisingly, higher concentrations of volatile acids (acetic acid) has a
negative impact on quality. This makes sense since high concentrations of acetic
acid give off a "vinegar" taste and aroma. 



```{r echo=FALSE}
# fixed acidity vs density
d1 <- ggplot(aes(x=density, y=fixed.acidity), data=df) +
  geom_jitter(alpha=0.3) +
  geom_smooth(method = 'lm') +
  ggtitle('Fixed acidity') 

# alcohol vs density
d2 <- ggplot(aes(x=density, y=alcohol), data = df) +
  geom_jitter(alpha=0.3) +
  geom_smooth(method = 'lm', color = 'red') +
  ggtitle('Alcohol') 

grid.arrange(d1, d2, heights=1) 

```

####
It's interesting to see how how the acids contribute to density. Fixed 
acidity, which is essentially a measure of all naturally occruing acids 
contained in wine makes up a large proportion of the body. In contrast, the 
higher the alcohol content the less dense the wine. This challenges my own
assumptions as I associate denser full bodied wines as less acidic. In this 
collection, at least, this is not the case. 


```{r echo=FALSE}
# pH vs fixed acidity
ggplot(aes(x=pH, y=fixed.acidity), data = df) +
  geom_jitter(alpha=0.25) +
  geom_smooth(method = 'lm', color = 'red') +
  ggtitle('Fixed acidity leads to wine to be even more acidic')
  theme_bw()
```

####
Not surprisingly, though to a small range, high fixed acidity levels lead to
an even lower pH.  Again, this explains whey wines can be tart, as acidic 
liquids consumed tend to be sour or tart. 



# Bivariate Analysis

####
When comparing quality, alcohol, citric acid, and sulphates emerged as the 
three attributes with the strongest positive correlation. I generated scatter
plots (with some jitter to spread the quality points) and added a trendline 
to visualize the magnitude of an effect. In contrast, the attribute with the 
largest negative correlation was volatile acid content. 

Another interesting relationship was comparing density with fixed acidity and 
alcohol, which gave strong, though opposite correlations. Here we found that 
higher acidity led to more denser wines. However, higher alcohol content 
denser wines would be due to presence of more alcohol and more acidic wines 
being less dense.  Going futher, pH was more acidic the higher the fixed acid
concentration. This finding also explains why pH has a moderate effect (-0.34)
on higher density wines. 

Overall the strongest positive correlations occurred between the following:
  -fixed acidity vs citric acid (R = 0.67)
  -fixed acidity vs density ( R= 0.67)
  -total SO2 vs free SO2 (R = 0.67)
  -pH vs fixed acidity (R = -0.68)
  -density vs alcohol (R = -0.50)



# Multivariate Plots Section

####
I wanted to further explore the relationship between acidity, alcohol, and 
their impact on quality. The majority of the dataset contained quality ratings
between 5-6 on a scale of 0-10, implying that these wines are mostly average. I
decided to subset the data to only show wines with a quality above 6.  The issue
here is that sample size is drastically reduced (217 rows). 

```{r echo=FALSE, Multivariate_Plots}
# fixed acidity vs alcohol at each rating 
hq1 <- ggplot(aes(x=alcohol, y=fixed.acidity, color=factor(df$quality))
              , data = df) +
  geom_point(size=3) +
  geom_smooth(method = 'lm', color='red') +
  scale_x_continuous(breaks = seq(9, 15, 1)) +
  scale_y_continuous(breaks = seq(6, 16, 2.5)) +
  facet_wrap(~quality) +
  scale_color_brewer(type = 'seq')

hq2 <- ggplot(aes(x=alcohol, y=fixed.acidity, color=factor(df$quality)), 
                  data = df) +
  geom_point(size=3) +
  geom_smooth(method = 'lm', color='red') +
  scale_x_continuous(breaks = seq(9, 15, 1)) +
  scale_y_continuous(breaks = seq(6, 16, 2.5)) +
  scale_color_brewer(type='seq')

grid.arrange(hq2, hq1)

```


####
Higher quality wines tend to have higher alcohol content. With regards to 
fixed acidity, we can see how the trendline changes trajectory downward when 
comparing the higher quality wines. Again, this plot removed a lot of points,
essentially focusing on a specific range of the output signal. 
Focusing back on the plot with all quality ratings, you can see that color of 
the plot gets lighter as alcohol at th very least increases to an industry mean 
range of 10-11%. If anything can be said, wines with less alcohol content less 
than 10% are likely to receive a  low quality rating. 


```{r echo=FALSE}
# volatile and citric acids vs quality
ggplot(aes(x = volatile.acidity, y = citric.acid, color=factor(df$quality)),
       data = df) + 
  geom_density2d(bins=3) + 
  scale_color_brewer(type = 'seq') + 
  geom_point(color='black',alpha=0.1) 

```

####
Since the other two acid attributes did effect quality of wine, I wanted to 
visualize citric acid vs volatile acid and wrap them in a cluster of specific
quality. Here we can see that, due to the most of the darker colored clusters 
occuring on the left hand side of the plot, citric acid with minimal volatile 
acidity concentrations leads to better quality wines. 

```{r echo=FALSE}
# citric acid and alcohol vs quality
ha1 <- ggplot(aes(x=alcohol,  y=citric.acid, color=factor(df$quality))
              , data =df) +
  geom_point(size=3) +
  geom_smooth(method = 'lm', color='red') +
  ylim(0.05, 1) +
  scale_color_brewer(type = 'seq')
  
# sulphates and alcohol vs quality
ha2 <- ggplot(aes(x=alcohol,  y=sulphates, color=factor(df$quality)),
              data = df)+
  geom_point(size=3) +
  geom_smooth(method = 'lm', color='red') +
  scale_color_brewer(type = 'seq')

grid.arrange(ha1, ha2)
  
```

####
Looking at the correlation table, I found it interesting that the only 
attributes that correlated postively with alcohol were citric acid and 
sulphates. Both of the additives have weak correlations with alcohol with citric
acid at 0.11, and  sulphates at 0.09. Since higher alcohol content led to higher
quality wine, I wanted to explore the effect against the additives. It's evident 
looking at the plots that the ligher colored points (higher quality) occur 
at higher concentrations of specific additive, but also alcohol. I noticed that 
the citric acid plot showed a density of points present at 0 g/dm3 so I removed 
them for the visualization by applying ylim on the y-axis. 



# Multivariate Analysis


####
Comparing fixed acidity and alcohol and their relationship with quality was
interesting even though the fixed acidity correlation is relatively weak 
(R=0.12). However, acidity has an postive correlation on density, while alcohol 
has a strong negative impact on density. If we're looking at the effects on 
quality, alcohol content is important to higher quality. Density, and by 
extension, the acidic components in a wine do not. 

Further, the non-naturally occuring attributes introduced into wine, citric 
acid and sulphates, have a positive effect on good quality wines. This may be
due to thier small effect to be associated with wines containing higher 
alcohol content. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# fixed acidity vs density
d3 <- ggplot(aes(x=density, y=fixed.acidity), data=df) +
  geom_jitter(alpha=0.3) +
  geom_smooth(method = 'lm') +
  theme_bw() +
  labs(x='Density (g/cm3)', 
       y='Fixed acidity (g/dm3)')

# alcohol vs density
d4 <- ggplot(aes(x=density, y=alcohol), data = df) +
  geom_jitter(alpha=0.3) +
  geom_smooth(method = 'lm', color = 'red') +
  theme_bw() +
  labs(x='Density (g/cm3)', 
        y='Alcohol (% by volume)')


grid.arrange(d3, d4, heights=1, top='The effects on Density') 

```

### Description One
This plot really shows how the relationship between acidity, alcohol, and 
density of the wine. Fixed acidity is increased due to the grapes and their 
skin, thus, giving a heavier profile.  Increasing alcohol content made the wine
less dense, and likely, of a lighter profile. 

### Plot Two
```{r echo=FALSE, Plot_Two}
# fixed acidity and alcohol vs quality at each rating
ggplot(aes(x=alcohol, y=fixed.acidity, color=factor(df$quality)), data = df) +
  geom_point(size=3) +
  geom_smooth(method = 'lm', color='red', se=FALSE) +
  scale_x_continuous(breaks = seq(9, 15, 1)) +
  scale_y_continuous(breaks = seq(6, 16, 2.5)) +
  facet_wrap(~quality) +
  scale_color_brewer(type = 'seq') +
  theme_light() +
  labs(x= 'Alcohol (% by volume)',
       y='Fixed acidity (g/dm3)', 
       title='Alcohol and Fixed acidity breakdown by quality rating')

```

### Description Two
I feel that this plot really shows how much alcohol affects quality. You can 
see at quality 3 and 4 that most of the points are at the left side of the 
plots, meaning lowere alcohol content. Contrast that to the wines with ratings
of 7-8 where the points are more prevalent to the center and right side of the 
x-axis, meaning higher alcohol. Similarly, we can see how acidity and wine have 
an almost inverse relationship with themselves, but also how fixed acidity 
essentially has not effect on quality. 

### Plot Three
```{r echo=FALSE, Plot_Three}
# citric acid and alcohol vs quality
ha1 <- ggplot(aes(x=alcohol,  y=citric.acid, color=factor(df$quality)), 
                  data =df) +
  geom_point(size=3) +
  geom_smooth(method = 'lm', color='red') +
  ylim(0.05, 1) +
  scale_color_brewer(type='seq') +
  labs(x='Alcohol (% by volumne)',
    y='Citric Acid (g/dm3)',
       title='Both citric acid and sulphate lead to better quality wines')

# sulphates and alcohol vs quality
ha2 <- ggplot(aes(x=alcohol,  y=sulphates, color=factor(df$quality)), 
              data = df) +
  geom_point(size=3) +
  geom_smooth(method = 'lm', color='red') +
  scale_color_brewer(type='seq') +
  labs(x='Alcohol (% by volumne)',y='Sulphates (g/dm3)')

grid.arrange(ha1, ha2 ) 
  
```

### Description Three
At this point we knew that alcohol was the most important factor in higher 
quality wine, but wanted to see if the additives have a positive effect on 
increasing alcohol content, thus, improving wine quality. It's hard to tell if 
these additives improved quality by modifying the profile and aroma of wine, 
or if they also strengthened the alcohol content. 


# Reflection

So upon completing the project, I realized I had a tough time getting started
due to a lack of domain knowledge. The data set itself was really tidy and easy
to work with, but without knowing what attributes do what, and what the data is 
trying to say, it's hard to find a story to tell. After the univariate analysis, 
I took a lot of time reading about wine chemistry to understand how various
attributes effect wine quality. It was interesting to see how additives (citric
acid and sulphates) used to add freshness and protect from further microbial 
growth, actually led to better quality wine. However, seeing that they also are 
correlated positively with more alcohol content in a wine, we question whether
the additives themselves and their ability to improve a wines taste and aroma 
are the cause of better wine. It was interesting project, and a reminder that,
while learning how to work with powerful analytical tools such as R are 
important, having subject knowledge on the topic simply cannot be overlooked 
exploring a problem. 

I think further analysis can be one a wine with a wider quality rating range, as
when you really consider these wines, the majority of them are average and makes
it difficult to really tease out what really makes a stand out wine. Further,
exploring whether a specific blend of attributes can lead to a quality wine
would be an interesting problem to tackle. 


References:

https://www.homebrewit.com/a-detailed-explanation-of-acids-used-in-wine-making

https://winefolly.com/wine-basics-beginners-guide/

https://www.rstudio.com/resources/cheatsheets/

https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt

